Manuscript draft

### 2019-05-31 manuscript/manuscript draft.Rmd
updated manuscript

Globally, forests play an important role in the carbon cycle,

@@

@@

*Understanding components of primary productivity.* Gross primary productivity (GPP) represents the gross uptake of carbon dioxide by plants in an ecosystem, to be used in photosynthesis and energy production. Only a fraction of this carbon is assimilated into plant tissues; the rest is used in autotrophic respiration. The component of GPP that is stored as plant material is the net primary productivity (NPP) of an ecosystem; thus GPP can be described as the sum of NPP and ecosystem respiration.

@@

*GPP* Gross primary productivity is generally thought to be highest in the tropics. Modelling of global terrestrial ecosystem GPP though upscaling and calibration of eddy flux measurements indicates a peak in GPP in the tropics, with the highest levels in tropical forests [@Beer2010; @Jung2011]. This is corroborated by analysis of site-level GPP measurements, which show a strong positive correlation between GPP and mean annual temperature (MAT) and mean annual precipitation (MAP) [@Luyssaert2007], with the highest GPP values reported in tropical forests. There is evidence that this increase is not linear: increases in GPP have been reported to saturate above 1500mm MAP [@Luyssaert2007] and 25$^\circ$C MAT [@Larjavaara2012].

*NPP* The influence of latitude on global patterns of NPP is less clear than that of GPP. Simova and Storch (2017) found that, as with GPP, NPP decreases with latitude, peaking in the tropics. However, other studies have found the highest values of NPP in temperate forests [@Huston2009], with increases in NPP saturating at 1500mm MAP and 10$^\circ$C MAT [@Luyssaert2007].

Because of the challenges in accurately obtaining unbiased measures of belowground productivity, many studies focus on ANPP in preference to measures of NPP. Studies on global patterns of ANPP are equally inconclusive: Gillman et al. (2015) found a weak negative relationship between ANPP and latitude, with the relationship becoming stronger in older forest stands. These findings were echoed in other studies, which have found weak or no relationships between ANPP and latitude [@Huston2009].

Primary productivity of forests is influenced by a range of factors. On a local scale, stand age [@Litton2007; @Gillman2015], management [@Simova2017]; nutrient availability [cite]; and altitude [@Girardin2010; @Malhi2017] all impact stand productivity. On a global scale, changes are likely to be influenced by climatic variables. The majority of studies to date have focused on the influence of MAP and MAT on productivity. While MAP and MAT have been argued to be significant predictors of productivity [@Chu2016], other studies have found that the correlation between productivity and MAT is a factor of the relationship between productivity and growing season length [@Kerkhoff2005; @Malhi2012; @Michaletz2014; @Michaletz2018].

>>>>

Globally, forests play an important role in the carbon cycle.

@@

The productivity of forests shows variation with latitude, however the exact nature of this pattern and the mechanisms that drive it are poorly understood. On a global scale variation in primary productivity is likely to be climate-driven, meaning that achieving a better understanding of the climatic controls on forest productivity is essential for understanding and predicting the reponses and feedbacks of forests to climate change.

@@

In order to better understand and predict responses and feedbacks to climate change, we need a better fundamental understand of how forest carbon cycling varies with respect to climate globally.

*Patterns of primary productivity.* Understanding variation in primary productivity is complex. Firstly, there are many factors which can influence productivity on a range of scales, many of which show interactive effects with each other. In addition, the primary productivity of a forest is the sum of the primary productivity of a number of individual components, which may all show individual responses to drivers of productivity. Understanding these relationships and interdependencies is complex.

Primary productivity of forests is influenced by a range of factors. On a local scale, stand age [@Litton2007; @Gillman2015], management [@Simova2017]; nutrient availability [cite]; and altitude [@Girardin2010; @Malhi2017] all impact stand productivity. On a global scale, changes in primary productivity are influenced by climatic variables and abiotic gradients, such as the length of growing season [@Michaletz2014]. The majority of studies to date have focused on the influence of MAP and MAT on productivity. While MAP and MAT have been argued to be significant predictors of productivity [@Chu2016], other studies have found that the correlation between productivity and MAT is a factor of the relationship between productivity and growing season length [@Kerkhoff2005; @Malhi2012; @Michaletz2014; @Michaletz2018]. The influence of other climate variables on global patterns of productivity remains under-explored.

*Understanding components of primary productivity.* There are several different measures of primary productivity. Gross primary productivity (GPP) represents the gross uptake of carbon dioxide by plants in an ecosystem, to be used in photosynthesis and energy production. Only a fraction of this carbon is assimilated into plant tissues; the rest is used in autotrophic respiration. The component of GPP that is stored as plant material is the net primary productivity (NPP) of an ecosystem; thus GPP can be described as the sum of NPP and ecosystem respiration.

@@

Current knowledge indicates that these different components of productivity show different latitudinal patterns. Gross primary productivity is generally thought to be highest in the tropics. Modelling of global terrestrial ecosystem GPP though upscaling and calibration of eddy flux measurements indicates a peak in GPP in the tropics, with the highest levels in tropical forests [@Beer2010; @Jung2011]. This is corroborated by analysis of site-level GPP measurements, which show a strong positive correlation between GPP and mean annual temperature (MAT) and mean annual precipitation (MAP) [@Luyssaert2007], with the highest GPP values reported in tropical forests. The influence of latitude on global patterns of NPP is less clear than that of GPP. Simova and Storch (2017) found that, as with GPP, NPP decreases with latitude, peaking in the tropics. However, other studies have found the highest values of NPP in temperate forests [@Luyssaert2007; @Huston2009]. Because of the challenges in accurately obtaining unbiased measures of belowground productivity, many studies focus on ANPP in preference to measures of NPP. Studies on global patterns of ANPP are equally inconclusive: Gillman et al. (2015) found a weak negative relationship between ANPP and latitude, with the relationship becoming stronger in older forest stands. These findings were echoed in other studies, which have found weak or no relationships between ANPP and latitude [@Huston2009].

These variations in the responses of different components of productivity to changes in latitude suggest that components may vary in the strength of their response to climatic drivers. There is some empirical evidence for this: increases in GPP are reported to saturate above 25$^\circ$C MAT [@Larjavaara2012], while increases in NPP are recorded to saturate above 10$^\circ$C MAT [@Luyssaert2007]. Similarly, allocation to different components of ANPP varies with climate. Within the tropics, allocation to canopy NPP appears fairly consistent, with significantly greater variation in allocation to woody and belowground NPP [@Litton2007; @Malhi2011]. Allocation to these structural biomass components has been shown to increase with water availability [@Litton2007; @Bloom2016], and is highest in the wet tropics, indicating that control of woody productivity by MAP may be more significant than control of foliar productivity. However, these studies are regional, meaning that our understanding of variation in allocation and its relationship to climate on a global scale remains limited.

Allocation is also influenced by stand age [@DeLucia2007], nutrient availability [@Litton2007] and forest structure [@Taylor2019], which can make it challenging to disentangle the effects of climate.
